Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
working on the model

%% MODEL
\newcommand{\effdims}[2]{\mathit{dims}(#1)_{#2}}

@@

\begin{definition}[Neighbourhood index and neighbourhood subscripts]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is a
  constant of integer type. That is, a neighbourhood index is a
  relative index defined as a constant translation of a base induction
  variable. The relation $\equiv$ identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ (\eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
  the predicate \neigh{}. An array subscript $v(\bar{e})$ where every
 $e \in \bar{e}$ is a neighbourhood index is called a
 \emph{neighbourhood subscript}.
\end{definition}

@@

\newcommand{\relix}{(\mathbb{Z}_\bot)^\mathbb{D}}

We define a semantic model of our stencil specification language as
the basis for building correct implementations, showing the soundness
of the equational theory and approximations, and as a basis for the
inference, checking, and program synthesis algorithms.
The semantics of stencil specifications can be given a denotational
model as sets of unifiers for array indexing.

\begin{definition}
  A specification of maximum dimensionality $n$, is modelled by a
  set of unifiers. Here, a unifier is represented as a vector of size $n$
  whose values are drawn from $\mathbb{Z} \cup \{\infty\}$. The purpose
  of the unifier is to accept or reject relativised indices of an array
  subscript. Formal definition of unification in this context is given
  later.

  The element $\infty$ represents the general acceptor meaning it would unify
  with any index in the same dimension including constants or any offset from
  an induction variable.
\end{definition}

\begin{definition} Unification relation, $\sim$, establishes whether
  a particular relativised subscript, $u$, of a group of subscripts involved
  in a stencil computation conforms with a unifier, $v$, of the model for
  the stencil computation.

  Formally, the unification relation can be defined as follows:
  $$u \sim_n v \triangleq
  \{ (u,v) \; \mid \; \forall i \in \{ 1, \ldots, n \}, v_i = \infty \vee u_i = v_i \}$$
\end{definition}

For example, consider a unifier of the form $\vtwoh{\infty}{1}$.

An array index term $(i, j+1)$ would corresponds to the vector
$\vtwoh{0}{1}$ after relativisation. This would unify with the
$\vtwoh{\infty}{1}$ as the first dimension of the unifier have a
general acceptor and the second dimension matches the relative index.

Another index term $(3, j)$ corresponds to the vector $\vtwoh{\infty}{0}$
after relativisation since $3$ is a non-relative index. This would fail
to unify. In the first dimension non-relative index would unify with
the general acceptor but in the second dimension 1 wouldn't unify with 3.

We define the interpretation function $\interp{-}_n$ where $n$ is the
maximum dimensionality of the specification being modelled, overloaded
on regions in Figure~\ref{fig:region-model} and on the top-level of a
specification \textit{spec} in Figure~\ref{fig:spatial-model}. Some
intermediate notations and definitions are used.

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector where the $r^{th}$ entry is $1$
and all others are $\infty$, \eg{}, $\singleEntry{1}{2} =
\vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.

Here the meaning of a scaled single-entry vector is that at the
dimension it is not $\infty$, it provides a unifier for consistent
relativised array subscript indices. In particular the constrained
requires a non-zero offset from the induction variable.
\end{definition}

@@

Similar to the single-entry vector, zero-entry vector also imposes
a constaint on relativised indices. The only difference is the indexing
expression crosses the origin with respect to the induction variable.
\end{definition}

The first four equations of Figure~\ref{fig:region-model}
give the meanings of the region constants as sets of unifiers.
The next two shows how regions are composed in the model.

\begin{definition}The product of two models of dimensionality
$n$ is given by the tensor $\otimes_n$ defined:
%%
\begin{alignat*}{2}
& \effdims{M}{n} =
\bigcup_{u \in M} \{x \mid x \in \{1,\ldots,n\}, u_x \neq \infty\} \\
& M \otimes_n N =
  \Bigg\{x \; \Bigg| \;
    \parbox{5.6cm}{$i \in \{1, \ldots, 2^n\},$
                   $(u, v) \in M \times N,$ \\
                   $x = (u \bowtie v)_i,$ \\
                   $j \in (\effdims{M}{n} \cup \effdims{N}{n}),$ \\
                   $x_j \neq \infty$
                  } \Bigg\}
\end{alignat*}
In plain language, tensor product combines non-$\infty$ constraints
of the unifiers in different dimensions into less accepting unifiers.
This means an access pattern unifying with this unifier would require
both models to be satisfied. Use of $dims$ gives the dimensions that
constitutent models talk about and is used in the product to eliminate
general unifier appearing in these dimensions. Allowing general unifier
in these dimensions would prevent both models to be satisfied
simultaneously.

The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
product of $M \times N$.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-deterministic picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise permutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$.
%%
\end{definition}

@@

\begin{theorem}[Soundness]
\[
\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
\; \Rightarrow \;
\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
\]
\end{theorem}

\paragraph{Proof} (see Appendix)

\begin{theorem}[Soundness]
\[
\overline{\texttt{v} : S} <: \overline{\texttt{u} : T}
\; \Rightarrow \;
\interp{\overline{\texttt{v} : S}} \subseteq \interp{\overline{\texttt{u} : T}}
\]
\end{theorem}

\paragraph{Proof} (see Appendix)

@@

\paragraph{Span coalescing, formally}

@@

The algorithm to create contiguous spans, covering the space of
indices, then proceeds as follows.
Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
created where each vector is mapped to the trivial \emph{unit span} is
created by pairing a vector with itself:
%
\begin{align*}
N(v) = \{[\vect{x} \;
  \vect{x}] \mid \vect{x} \in M(v)\}
\end{align*}
%
For our example, this yields:
%%
\begin{align*}
N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
\end{align*}
%%
A fixed point is then compute for the following
procedure \textsf{regions} which, for per variable
in the map, coalesces spans into contiguous regions in
the $n$-dimensional space. That is, we compute
 $(\mu \; \textsf{regions}) N$ where \textsf{regions}
is defined as follows for each variable in the map:
%
\begin{enumerate}

@@

\item Sort each permutation set into an ordered list, based on the ordering:
\begin{equation*}
  \vect{x} \leq \vect{y} =
      \exists i \, . \; \vect{x}_{i,1} \leq \vect{y}_{i,1} \; \wedge \;
        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
\end{equation*}
%
This yields:
%%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%%
\item Fold each list pairwise by the following partial operation
 $\bullet$ which coalesces contiguous regions:
%
\begin{align*}
& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
\vect{x}_{2:n} = \vect{y}_{2:n}) \\
\bot  & \textit{otherwise}
\end{cases}
\end{align*}
For our example, this yields:
%
\begin{align*}
Q(v) = \{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}
]) \\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\item Un-permute and union together, \ie{},
%
\[
U(v) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
 \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
\]
Which yields:
%
\begin{align*}
U(\texttt{a}) =
\{\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{-1}{0}{1},
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\item Finally, filter by the region containment predicate $\containedin$, that
  is, if any region is contained within another then remove the
  smaller. The $\containedin$ predicate is defined:
%
\begin{align*}
& (\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = \\
& \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
\end{align*}
For our example this then yields the final result of:
\begin{align*}
\textsf{regions}(M)
= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
\end{align*}
since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
\end{enumerate}
For our example, applying \textsf{regions} again yields the same
result, hence we have reached a fixed point.

@@

\newcommand{\bplus}{\operatornamewithlimits{\term{+}}}
\newcommand{\tySum}[1]{#1^{\term{+}}}
\newcommand{\tyProd}[1]{#1^{\term{*}}}
\newcommand{\specDNF}{\textbf{spec}}

Given the set $\textsf{regions}(M)$ of spans covering the indexing
space, the next step converts these into abstract syntax trees
for the specification language. This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}). The algebra mirrors the shape
of the region syntax with additive and multiplicative binary operators
$\oplus$ and $\odot$ and a constructor for \specDNF{}:
%%
\begin{align*}
\oplus & : \specDNF \rightarrow \specDNF \rightarrow
  \specDNF \\
\odot & : \specDNF \rightarrow \specDNF \rightarrow
        \specDNF \\
\mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
  \{\infty\})  \rightarrow \specDNF
\end{align*}
%%
The $\mathsf{toSpec}$ operation takes a pair of lower and upper bounds
values (one-dimensional) and constructs a specification (in the
\textit{spec} syntax) which contains a one-dimensional
region. An entire span is convereted then into a $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of bounds in a span, per dimension, and multiplying these
specifications by $\odot$. The resulting specifications for
each span are then summed by by $\oplus$, that is:
\begin{equation*}
\textit{spec}(M) = 
\sum\limits_{\vect{x} \in \textsf{regions}(M)}
\prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
(where $\oplus$ is the additive and $\odot$ is the multiplicative used
within $\Sigma$ and $\Pi$ respectively here).

\begin{definition}[DNF specifications]
Let $r_1, ..., r_n$ range over elements of the \textit{region}
syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax})
The domain of \specDNF{} specifications
in disjunctive normal form ranges over terms of the form:
%%
(r_1 \term{*} \ldots \term{*} r_n)\term{+} \ldots 
\term{+} (r_1 \term{*} \ldots \term{*} r_n)
%%
\end{definition}

>>>>

%% MODEL
\newcommand{\effdims}[2]{\mathit{constr}(#1)_{#2}}

@@

\begin{definition}[Neighbourhood index and neighbourhood subscripts]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is a
  constant of integer type. That is, a neighbourhood index is a
  relative index defined as a constant translation of a base induction
  variable. The relation $\equiv$ identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ (\eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
  the predicate \neigh{}. An array subscript $v(\bar{e})$ where every
 $e \in \bar{e}$ is a neighbourhood index is called a
 \emph{neighbourhood subscript}.
\label{def:neighbour}
\end{definition}

@@

\newcommand{\domainVal}{\mathbb{Z}_\infty}

We define a semantic model for our stencil specification language.
This model has a number of purposes: (1) it is used in the inference,
checking, and program synthesis algorithms (introduced in
Section~\ref{sec:analysis}); (2) gives a basis for the equational
theory and approximations of the previous section; and (3) can be used
to guide correct implementations. The model is a denotational
semantics, in terms of sets of vectors which we call \emph{index
  schemes}. We introduce first the notion of indexing schemes
and \emph{unification} which gives the domain of our model
and forms the basis of the rest of the algorithsm for our
specificaiton. We then define the denotational model

\subsection{Indexing schemes and unification}

\begin{definition}[Index scheme]
An \emph{index scheme} is a vector of size $n$ (an $n$-vector)
with values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$. 
Values in $\mathbb{Z}$ represent offsets 
of a neighbourhood index (indexing
expressions of the form $i + a$ where $i$ is an induction variable
and $a$ is a constant offset, Definition~\ref{def:neighbour},
p.~\pageref{def:neighbour}). The additional value
$\infty$ represents any indexing behaviour--- it is a 
wildcard.
Throughout $u, v, w$ will range over index schemes.
\end{definition}

Indexing schemes give an abstract
representation of index expressions in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours allowed under the specification.
The relationship between indexing schemes and source language
terms is given precisely by the partial function, \textsf{schematic}, which
maps syntactic array index terms to indexing schemes. This
operation forms the basis of the analysis required for the inference
and the checking procedures in Section~\ref{sec:analysis}.

\begin{definition}[Schematic]
The partial function $\textsf{schematic}$ maps 
source-language indexing terms $(e_1, \ldots, e_n)$
into $n$-dimensional indexing schemes 
($\in \domainVal^n$), defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & = 
[\downarrow\!e_1 \ldots \downarrow\!e_n ] 
\quad \textit{iff} \; \forall i . \downarrow\!e_i \neq \bot 
\end{align*}
%
The $\downarrow$ operator is a partial\footnote{We represent
  partiality explicitly in our types with the set $A + \bot$.} function 
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
which maps each individal indexing expression to a component of
the indexing scheme, defined: 
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
appearing in an expression, and $\neigh{}$ classifies expressions which 
are constant offsets from an induction variable
(Defn.~\ref{def:neighbour}). In the third case, 
if the expression is not a neighbourhood index but is not
a relative index either (\ie{} is not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

For example $\textsf{schematic}(i, j+1) = \vtwoh{0}{1}$ assuming
$i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
 = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
 example, is
 undefined as its first index is a non-neighbour relative index, which 
 is outside the scope of our specification language.
\end{definition}

\begin{definition}[Unification]
Two schemes $u, v$ of dimensionality $n$
are said to be \emph{unifiable}, written
$u \sim_n v$, iff the following holds:
%%
\[
u \sim_n v \; \Leftrightarrow  \;
  \forall i\!\in\!\{ 1,.., n \} . \, (u_i = \infty
\, \vee \, v_i = \infty \, \vee \, u_i = v_i)
\]
%%
That is, two schemes are related by the binary relation
$\sim_n$ over $(\mathbb{Z}_{\infty})^n$ if in each dimension
either is $\infty$ or they are equal. 
Thus, the element $\infty$ represents the ``general acceptor''
 meaning it unifies with any index in the same dimension.

The subscript on $\sim_n$ is sometimes dropped when the
intended dimensionality is clear or inconsequential.
\end{definition}

%Unification relation, $\sim$, establishes whether
%a particular relativised subscript, $u$, of a group of subscripts involved
%in a stencil computation conforms with a unifier, $v$, of the model for
%the stencil computation.

\begin{example}
Take three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$.
As given by \textsf{schematic}, these have the 
corresponding schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationship of unification (or lack of) then holds between
them:
%%
\begin{equation*}
\vtwoh{\infty}{0} \; \not\sim_2 \; \vtwoh{0}{1} \; \sim_2 \;
\vtwoh{\infty}{1} \; \not\sim_2 \; \vtwoh{\infty}{0}
\end{equation*}
%%
The middle pair unifies as $0$ and $\infty$ unify, 
whereas the left and right pairs do not unify as
$0 \not\sim_1 1$. 
\end{example}

Our terminology of \emph{indexing schemes} and
\emph{unification} are by analogy
with the \emph{type schemes} and \emph{type unification} of
ML~\cite{milner1978theory}, though our notion is much more rigid.

\subsection{Denotational model}

The model is given by the interpretation function $\interp{-}_n$
(where $n$ is the maximum dimensionality of the specification being
modelled) mapping closed\footnote{That is, we assume that there are no
  occurences of \textit{rvar} in a specification being modelled.  Any
  \emph{open} specification containing region variables can be made
  closed by straightforward syntactic subtitution with a (closed) 
  \textit{region}.} specifications to sets of
$n$-dimensional indexing schemes. The interperation is overloaded on
\emph{regions} in Figure~\ref{fig:region-model} and on the top-level
of a specification \textit{spec} in Figure~\ref{fig:spatial-model}.

We first define some intermediate notations and definitions.

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector where the $r^{th}$ entry is $1$
and all others are $\infty$, \eg{}, $\singleEntry{1}{2} =
\vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
Thus, a single-entry vector is an indexing scheme which unifies with
all other schemes which have $1$ as their $r^{th}$ entry,
describing a neighbour offset of $1$. 
%DAO don't understand this sentence:
%In particular the constrained
%requires a non-zero offset from the induction variable.
\end{definition}

@@

Similarly to the single-entry vector, zero-entry vectors 
are also indexing schemes which unify with all other schemes
as long as the $r^{th}$ entry is $0$, representing indices at
the ``origin'' in dimension $r$. 
\end{definition}

\noindent
The first four equations of Figure~\ref{fig:region-model}
give the meanings of the region constants as sets of indexing
schemes. The next two give the meaning of combining regions
by $\term{+}$ and $\term{*}$.

\begin{example}
For a $2$-dimensional situation, 
\begin{align*}
& \interp{\stenCenS{2}{1}}_2 \\
= \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\
= \; & \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\
= \; & \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty}, 
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\} 
\end{align*}
Note the absboring behaviour of $\infty$ with respect to
multiplication. 
\end{example}
\noindent
The model of $\term{+}$ is the straightforward union of 
the models of subterms. The model of $\term{*}$ is uses
an intermediate tensor $\otimes$ which requires some intermediate
definitions.

\begin{definition}[Constrained dimensions]
Given an $n$-dimensional model $M$, the \emph{constrained dimensions}
of that model is a subset of $\{1, \ldots, n\}$ giving
the dimensions for which at least one scheme $u \in M$ has
a non-$\infty$ value in that dimension. That is:
%
\begin{align*}
\effdims{M}{n} =
\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
  \infty\}
\end{align*}
For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
 = \{1\}$.
\end{definition}

\begin{definition}The tensor $\otimes_n$ of $n$-dimensional
models is defined:
%
%\begin{align*}
%\{\vtwoh{0}{0}, \vtwoh{1,1}\}
% \otimes \{\vtwoh{0,\infty}\}
%\end{align*}
%
%
%%
\begin{alignat*}{2}
& M \otimes_n N =
  \Bigg\{x \; \Bigg| \;
    \parbox{6cm}{$(u, v) \in M \times N $ \\[0.1em]
                 $\;\; \wedge \;\, j \in (\effdims{M}{n} \cup
                   \effdims{N}{n})$\\[0.1em]
                  $\;\; \wedge \;\, i \in \{1, \ldots, 2^n\} \, \wedge
                  \, x = (u \bowtie v)_i$ \\[0.1em]                   
                   $\;\; \wedge \;\, x_j \neq \infty$
                  } \Bigg\} 
\end{alignat*}
The intuition for the tensor is that it gives all possible
ways of combining indexing schemes in two models
such that for every scheme
$u \in (M \otimes_n N)$ then $w \sim u$ for some $w$
\emph{iff} $\;\exists v \in M . w \sim v$ and
$\exists v' \in N . w \sim v'$. Thus it acts as a kind of intersection
for indexing schemes. We unpack its definition in pieces.

We start with the first guard taking the Cartesian product of
the two models and quantifying over all pairings $(u, v)$. 
The second guard says that $j$ is in the set of constrainted
dimensions for $M$ and the constrained dimensions for $N$.
That is, every $j$ corresponds to a dimension in which 
either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
and thus is constraining.

%for all possible
%pairs of schemes $(u, v)$ in the Cartesian product of the
%two models $M \times N$, then

% combines non-$\infty$ constraints
%of the unifiers in different dimensions into less accepting unifiers.
%This means an access pattern unifying with this unifier would require
%both models to be satisfied. Use of $dims$ gives the dimensions that
%constitutent models talk about and is used in the product to eliminate
%general unifier appearing in these dimensions. Allowing general unifier
%in these dimensions would prevent both models to be satisfied
%simultaneously.

The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors), which is indexed by $i$ above
and then $j$. Using this, the $\otimes$ operation constructs the set
of schemes which are the 
pairwise permutations of every pair $(u, v)$ of schemes in the Cartesian
product of $M \times N$. These permuted schemes are only those
which have non-$\infty$ components in the constrained dimensions
given by each model.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-deterministic picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise permutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$.
%%
\todo{Complete this}
\end{definition}

@@

\paragraph{Model of \textit{spec}}

\todo{Explain the injections and the domain here}

\subsection{Soundness}

Given the model defined in this section we can
consider soundness for the equational theory of
specifications and the notion of spatial approximation
on specifications. The proofs are given in our supplement.

\begin{theorem}[Soundness of equational theory]
\[
\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
\; \Rightarrow \;
\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
\]
\end{theorem}

\begin{theorem}[Soundness of approximations]
\[
\overline{\texttt{v} : S} <: \overline{\texttt{u} : T}
\; \Rightarrow \;
\interp{\overline{\texttt{v} : S}} \subseteq \interp{\overline{\texttt{u} : T}}
\]
\end{theorem}

@@

\newcommand{\spanOp}{\textsf{spans}}
\paragraph{Span coalescing, formally}

@@

The algorithm to create contiguous spans, covering the space of
indices, then proceeds as follows.
Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
created where each vector is mapped to the trivial \emph{unit span} is
created by pairing a vector with itself:
%
\begin{align*}
N(v) = \{[\vect{x} \;
  \vect{x}] \mid \vect{x} \in M(v)\}
\end{align*}
%
For our example, this yields:
%%
\begin{align*}
N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
\end{align*}
%%
A fixed point is then compute for the following
procedure \spanOp which, for per variable
in the map, coalesces spans into contiguous regions in
the $n$-dimensional space. That is, we compute
 $(\mu \; \spanOp) N$ where \spanOp
is defined as follows for each variable in the map:
%
\begin{enumerate}

@@

\item Sort each permutation set into an ordered list, based on the ordering:
\begin{equation*}
  \vect{x} \leq \vect{y} =
      \exists i \, . \; \vect{x}_{i,1} \leq \vect{y}_{i,1} \; \wedge \;
        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
\end{equation*}
%
This yields:
%%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%%
\item Fold each list pairwise by the following partial operation
 $\bullet$ which coalesces contiguous regions:
%
\begin{align*}
& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
\vect{x}_{2:n} = \vect{y}_{2:n}) \\
\bot  & \textit{otherwise}
\end{cases}
\end{align*}
For our example, this yields:
%
\begin{align*}
Q(v) = \{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}
]) \\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\item Un-permute and union together, \ie{},
%
\[
U(v) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
 \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
\]
Which yields:
%
\begin{align*}
U(\texttt{a}) =
\{\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{-1}{0}{1},
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\item Finally, filter by the region containment predicate $\containedin$, that
  is, if any region is contained within another then remove the
  smaller. The $\containedin$ predicate is defined:
%
\begin{align*}
& (\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = \\
& \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
\end{align*}
For our example this then yields the final result of:
\begin{align*}
\spanOp(M)
= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
\end{align*}
since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
\end{enumerate}
For our example, applying \spanOp again yields the same
result, hence we have reached a fixed point.

@@

\newcommand{\bplus}{\operatornamewithlimits{\term{+}}}
\newcommand{\tySum}[1]{#1^{\term{+}}}
\newcommand{\tyProd}[1]{#1^{\term{*}}}
\newcommand{\specDNF}{\textbf{spec}}
\newcommand{\result}[1]{\lfloor{#1}\rceil}

Next, the set of spans $\spanOp(M)$ covering the indexing
space is converted into abstract syntax trees 
for the specification language. This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}).

\begin{align*}
\result{\specDNF{}} = \specDNF + ((\specDNF{} + 1) \times
  (\specDNF{} + 1))
\end{align*}

\begin{align*}
\result{\specDNF{}} & ::= \textsf{exact} \, \specDNF{}
\mid \textsf{bounds} ([\specDNF{}], [\specDNF{}]) \\
\end{align*}

The algebra mirrors the shape
of the region combinators for addition and multiplication, 
with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
%%
\begin{align*}
\oplus & : \specDNF \rightarrow \specDNF \rightarrow
  \specDNF \\
\odot & : \specDNF \rightarrow \specDNF \rightarrow
        \specDNF \\
\mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
  \{\infty\})  \rightarrow \specDNF
\end{align*}
%%
The $\mathsf{toSpec}$ operation takes a pair of lower and upper bounds
values (one-dimensional) and constructs a specification 
which contains a one-dimensional region. 
An entire span is converted into an $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of lower and upper bounds in a span, per dimension, and multiplying these
specifications by $\odot$. The resulting specifications for
each span are then summed by by $\oplus$, that is, this step is given by:
\begin{equation*}
\textsf{spec}(M) = 
\sum\limits_{\vect{x} \in \spanOp(M)}
\prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
(where $\oplus$ is the additive and $\odot$ is the multiplicative used
within $\Sigma$ and $\Pi$ respectively here).
We unpack the definitions of this algebra.
